---
title: "Statistical analysis"
author: "Adrian Lison"
---

This notebook contains the main statistical analysis conducted in Föll, Lison, Maritsch et al. 2022.

# Preparation
```{r, message=F}
source("setup.R")
knitr::purl("preprocessing/preparation.Rmd", "preprocessing/preparation.R")
source("preprocessing/preparation.R")

source("utils/utils.R")
source("utils/utils_performance.R")
source("utils/utils_plot.R")
```

# Descriptives
Patients
```{r}
cat(paste("Overall cohort:", length(surv_data %>% pull(ID) %>% unique()) + length(exclude)))

cat(paste0(
  "\nExcluded: ", length(exclude), "\n  ",
  length(exclude_negative), " due to negative test\n  ",
  length(exclude_measurement), " due to nonadherence to the prescribed measurement regime\n  ",
  length(exclude_short_stay), " due to hospital discharge on the same day of hospitalization"
))

cat(paste("\nIncluded in analysis:", length(surv_data %>% filter(!(ID %in% exclude)) %>% pull(ID) %>% unique())))

cat("\n ", length(icu_patients), "ICU patients")
cat("\n ", length(surv_data %>% filter(!(ID %in% c(exclude, icu_patients, dropouts))) %>% pull(ID) %>% unique()), "Discharged patients")
cat("\n ", length(dropouts), "Censored drop-outs")
```
Observations
```{r}
# Original number of observations:
cat("Number of observations before quality checks:", surv_data %>%
  filter(!(ID %in% exclude)) %>%
  nrow())

# Number of observations after application of all quality criteria
cat("\nNumber of observations after quality checks:", surv_data %>%
  filter(!(ID %in% exclude)) %>%
  ensure_quality_observations(observation_coverage = observation_coverage, coverage_factor = coverage_factor) %>%
  nrow())
```
Demographics
```{r}
demographics %>%
  filter(ID %in% (surv_data %>%
    filter(if_any(c(HEART_RATE_mean, hrv_mean_hr), ~ !is.na(.))) %>%
    pull(ID) %>%
    unique())) %>%
  mutate(outcome = case_when(ID %in% icu_patients ~ "icu", ID %in% dropouts ~ "dropout", T ~ "dismissed")) %>%
  group_by(outcome) %>%
  summarize(across(age, list(mean = mean, sd = sd, min = min, max = max)),
    sex_male = sum(sex == "male"),
    sex_female = sum(sex == "female"),
    n = n()
  )
```

# Explanatory Analysis

## Physiological features
```{r}
statistical_feature_prefixes <- c("_max", "_mean", "_min", "_rms", "pct_95", "pct_5", "_iqr", "_iqr_5_95", "_std", "_entropy", "_n_above_mean", "_n_below_mean", "_energy")
hrv_feature_names <- c("hrv_mean_nni", "hrv_median_nni", "hrv_range_nni", "hrv_sdsd", "hrv_rmssd", "hrv_nni_50", "hrv_pnni_50", "hrv_nni_20", "hrv_pnni_20", "hrv_cvsd", "hrv_sdnn", "hrv_cvnni", "hrv_total_power", "hrv_vlf", "hrv_lf", "hrv_hf", "hrv_lf_hf_ratio", "hrv_lfnu", "hrv_hfnu")

predictors <- surv_data %>%
  select(
    starts_with(c("HEART_RATE", "RESPIRATION")) & ends_with(statistical_feature_prefixes),
    all_of(hrv_feature_names)
  ) %>%
  names()
```

## Model fitting
```{r}
for (predictor in predictors) {
  model_formula <- as.formula(paste("event ~ 1 + (1 | ID) + rel_day + age + sex +", predictor))

  model_data <- surv_data %>%
    ensure_quality_observations(predictor, observation_coverage, coverage_factor) %>%
    mutate(rel_day = scale(rel_day)) %>%
    drop_na(all.vars(model_formula))

  prior_univar <- prior(normal(0, 1), class = "b") +
    prior(normal(0, 5), class = "Intercept") +
    prior(normal(0, 1), class = "b", coef = "rel_day") +
    prior(student_t(3, 0, 1), class = "sd", group = "ID")

  model <- brm(model_formula,
    data = model_data,
    family = cumulative(link = "cloglog", threshold = "flexible"),
    prior = prior_univar,
    control = list(adapt_delta = 0.9),
    file = paste0("models/main/explanatory_", predictor),
    file_refit = "on_change"
  )
}
```

## Association with patient outcomes
Load fitted models and arrange data
```{r}
m_fits <- bind_rows(lapply(predictors, function(predictor) {
  if (file.exists(paste0("models/main/explanatory_", predictor, ".rds"))) {
    model <- readRDS(paste0("models/main/explanatory_", predictor, ".rds"))
    return(model %>% gather_draws(!!as.name(paste0("b_", predictor)), regex = T))
  } else {
    print(paste(predictor, "does not exist."))
    return(data.frame())
  }
})) %>%
  mutate(.variable = str_remove(.variable, "b_"))

m_fits_ranks_hybrid <- m_fits %>%
  group_by(.variable) %>%
  summarize(.value = quantile(.value, 0.1), .groups = "drop") %>%
  mutate(rank = min_rank(.value))

m_fits_plotdata <- m_fits %>%
  left_join(m_fits_ranks_hybrid %>% select(.variable, rank), by = ".variable") %>%
  mutate(measure_type = case_when(
    str_contains(.variable, "HEART_RATE_") ~ "HEART_RATE",
    str_contains(.variable, "RESPIRATION_") ~ "RESPIRATION",
    str_contains(.variable, "hrv_") ~ "HRV",
    str_contains(.variable, "num_ibis") ~ "HRV",
    T ~ "other"
  )) %>%
  ungroup() %>%
  mutate(.variable.name = str_remove(.variable, paste0(measure_type, "_")), .variable = reorder(.variable, rank, order = T)) %>%
  arrange(measure_type, .variable)
```

Figure 3
```{r,fig.height=5,fig.width=8}
plot_grid(
  plotlist = lapply(c("HEART_RATE", "HRV", "RESPIRATION"), get_univar_plot),
  ncol = 3, align = "v", labels = letters[1:3], label_fontfamily = "Times New Roman"
)

ggsave("figures/explanatory_coef.png", width = 15, height = 7.5, units = "cm", dpi = 300)
```

# Development of the Risk Score

## Selection of physiological features
```{r}
significant_features <- m_fits %>%
  group_by(.variable) %>%
  select(.variable, .value) %>%
  mean_qi(.width = 0.8) %>%
  filter(.lower > 0 | .upper < 0) %>%
  pull(.variable)

significant_features
```

## Principal component analysis (PCA)
Compute principal components (PCs)
```{r}
PCA_observations <- observation_coverage %>%
  filter(HEART_RATE_n_obs >= coverage_factor & RESPIRATION_n_obs >= coverage_factor & hrv_n_windows_notna >= coverage_factor) %>%
  select(ID, rel_day)
cl_pc <- prcomp(surv_data %>% inner_join(PCA_observations, by = c("ID", "rel_day")) %>% select(all_of(significant_features)) %>% drop_na(), scale. = TRUE)
summary(cl_pc)
```

Save PCs into data frame for further analysis
```{r}
surv_data_PCs <- surv_data %>%
  inner_join(PCA_observations, by = c("ID", "rel_day")) %>%
  select(ID, datetime, rel_day, event, age, sex, all_of(significant_features)) %>%
  drop_na() %>%
  select(ID, datetime, rel_day, event, age, sex) %>%
  bind_cols(data.frame(cl_pc$x)) %>%
  mutate(across(starts_with("PC"), scale)) %>%
  left_join(surv_data, by = c("ID", "rel_day", "event", "age", "sex"))
```

```{r}
saveRDS(surv_data_PCs, file = "../data/main/pca/data_pca.rds")
```

## LASSO feature selection

Load data from the PCA
```{r}
surv_data_PCs <- readRDS("../data/main/pca/data_pca.rds")
```

Select PC features via LASSO
```{r}
# Features for the main risk score
get_PCs_LASSO(surv_data_PCs)

# Features for the fixed risk score (no update of physiological data, see performance evaluation)
get_PCs_LASSO(
  surv_data_PCs %>%
    group_by(ID) %>%
    filter(min(rel_day) == 1) %>%
    mutate(across(starts_with("PC"), ~ .[1])) %>%
    ungroup()
)
```
## Computation of risk score
```{r}
risk_score_formula <- event ~ 1 + (1 | ID) + rel_day + age + sex + PC1 + PC11 + PC12 + PC18 + PC2 + PC20 + PC21 + PC3

prior_multivar <- prior(normal(0, 1), class = "b") +
  prior(normal(0, 5), class = "Intercept") +
  prior(normal(0, 1), class = "b", coef = "rel_day") +
  prior(student_t(3, 0, 1), class = "sd", group = "ID")

risk_score_observations <- observation_coverage %>%
  filter(HEART_RATE_n_obs >= coverage_factor & RESPIRATION_n_obs >= coverage_factor & hrv_n_windows_notna >= coverage_factor) %>%
  select(ID, rel_day)

risk_score_data <- surv_data_PCs %>%
  inner_join(risk_score_observations, by = c("ID", "rel_day")) %>%
  mutate(rel_day = scale(rel_day)) %>%
  drop_na(all.vars(risk_score_formula))

risk_score_model <- brm(
  risk_score_formula,
  data = risk_score_data,
  family = cumulative(link = "cloglog", threshold = "flexible"),
  chains = 4, inits = 0,
  prior = prior_multivar,
  file = "models/main/risk_score", file_refit = "on_change"
)
```

Figure 4
```{r}
risk_score_probs <- risk_score_model %>%
  spread_draws(b_Intercept[i]) %>%
  pivot_wider(.draw, names_from = i, values_from = b_Intercept, names_prefix = "tau") %>%
  mutate(ind = 1) %>%
  inner_join(data.frame(ind = 1, pred = seq(-3, 3, 0.1)), by = "ind") %>%
  select(-ind) %>%
  mutate(
    discharge_prob = discharge_prob(pred, tau1, tau2),
    stay_prob = stay_prob(pred, tau1, tau2),
    icu_prob = icu_prob(pred, tau1, tau2)
  ) %>%
  pivot_longer(ends_with("_prob"))

get_probability_plot(risk_score_probs)

ggsave("figures/risk_score_probs.png", height = 9.375, width = 15, units = "cm", dpi = 300)
```

Figure 7 (Multimedia Appendix)
```{r,fig.width=12,fig.height=4}
# plot coefficients
pc_lasso_effects_plot <- risk_score_model %>% get_multivar_plot()

# plot PCA loadings
PC_direction <- tribble(
  ~PC, ~direc,
  "PC1", 1,
  "PC11", 1,
  "PC12", -1,
  "PC18", -1,
  "PC2", 1,
  "PC20", 1,
  "PC21", 1,
  "PC3", 1,
)

pc_loadings_df <- data.frame(cl_pc$rotation) %>%
  rownames_to_column() %>%
  select(rowname, PC1, PC11, PC12, PC18, PC2, PC20, PC21, PC3) %>%
  pivot_longer(-rowname, names_to = "PC") %>%
  rename(feature = rowname) %>%
  mutate(measure_type = case_when(
    str_detect(feature, "HEART_RATE_") ~ "HEART_RATE",
    str_detect(feature, "RESPIRATION_") ~ "RESPIRATION",
    str_detect(feature, "hrv_") ~ "HRV",
    str_detect(feature, "num_ibis") ~ "HRV",
    T ~ "other"
  )) %>%
  mutate(feature.name = str_remove(feature, "HEART_RATE_|RESPIRATION_|hrv_")) %>%
  mutate(feature.name = factor(feature.name, ordered = T, levels = rev(c("max", "mean", "min", "rms", "pct_95", "pct_5", "energy", "iqr", "iqr_5_95", "std", "entropy", "perm_entropy", "svd_entropy", "n_above_mean", "n_below_mean", "mean_nni", "median_nni", "range_nni", "sdsd", "rmssd", "nni_50", "pnni_50", "nni_20", "pnni_20", "cvsd", "sdnn", "cvnni", "total_power", "vlf", "lf", "hf", "lf_hf_ratio", "lfnu", "hfnu")))) %>%
  left_join(PC_direction, by = "PC") %>%
  mutate(direction_influence = value * direc > 0) %>%
  mutate(PC = factor(PC, ordered = T, levels = PC_direction$PC)) %>%
  mutate(measure_type = str_replace(measure_type, "_", " "))

pc_loadings_plot <- get_loadings_plot(pc_loadings_df)

# Combine plots
plots <- align_plots(pc_lasso_effects_plot, pc_loadings_plot, align = "v", axis = "lr")
plot_grid(white_plot, plots[[1]], white_plot, plots[[2]], ncol = 2, rel_heights = c(0.6, 1), rel_widths = c(0.03, 0.97), labels = c("", "a", "", "b"))

ggsave("figures/risk_score_loadings.png", height = 13, width = 15, units = "cm", dpi = 300)
```

# Performance Evaluation

## Data and model preparation
```{r}
cv_observations <- observation_coverage %>%
  filter(HEART_RATE_n_obs >= coverage_factor & RESPIRATION_n_obs >= coverage_factor & hrv_n_windows_notna >= coverage_factor) %>%
  select(ID, rel_day)
```

Main risk score model (with updating)
```{r}
# Train
cv_data_update_train <- surv_data %>%
  inner_join(cv_observations, by = c("ID", "rel_day")) %>%
  drop_na(all_of(significant_features)) %>%
  arrange(ID, rel_day)

# Test
cv_data_update_test <- surv_data %>%
  drop_na(all_of(significant_features)) %>%
  arrange(ID, rel_day)

# Model
cv_model_update <- brm(
  formula = risk_score_model$formula,
  data = transform_pca(cv_data_update_train, list(cv_data_update_train), significant_features)[[1]],
  family = cumulative(link = "cloglog", threshold = "flexible"),
  chains = 4, inits = 0,
  prior = prior_multivar,
  file = "models/performance/cv_model_update", file_refit = "on_change"
)
```

Fixed risk score model (without updating)
```{r}
# Train
cv_data_noupdate_train <- cv_data_update_train %>%
  group_by(ID) %>%
  filter(min(rel_day) == 1) %>%
  mutate(across(all_of(significant_features), ~ .[1])) %>%
  ungroup()

# Test
cv_data_noupdate_test <- cv_data_update_test %>%
  group_by(ID) %>%
  mutate(across(all_of(significant_features), ~ .[1])) %>%
  ungroup()

# Model
cv_model_noupdate <- brm(
  formula = event ~ 1 + (1 | ID) + rel_day + age + sex + PC1 + PC11 + PC12 + PC13 + PC17 + PC18 + PC20 + PC21 + PC3,
  data = transform_pca(cv_data_noupdate_train, list(cv_data_noupdate_train), significant_features)[[1]],
  family = cumulative(link = "cloglog", threshold = "flexible"),
  chains = 4, inits = 0,
  prior = prior_multivar,
  file = "models/performance/cv_model_noupdate", file_refit = "on_change"
)
```

"NUll model" risk score with demographics only
```{r}
# Here, train and test data can be the same as for the main model (but only demographic features will be used by the null model)

# Model
cv_model_demographics_only <- brm(
  formula = event ~ 1 + (1 | ID) + rel_day + age + sex,
  data = transform_pca(cv_data_update_train, list(cv_data_update_train), significant_features)[[1]],
  family = cumulative(link = "cloglog", threshold = "flexible"),
  chains = 4, inits = 0,
  prior = prior(normal(0, 1), class = "b") + prior(normal(0, 5), class = "Intercept") + prior(normal(0, 1), class = "b", coef = "rel_day") + prior(normal(0, 0.1), class = "sd", group = "ID"),
  file = "models/performance/cv_model_demographics_only", file_refit = "on_change"
)
```

## Leave-One-Subject-Out CV

Main risk score model (with updating)
```{r}
kf_pred_update <- kfold_manual(
  cvmodel = cv_model_update,
  n_folds = length(unique(cv_data_update_test$ID)),
  pca_features = significant_features,
  data_training = cv_data_update_train,
  data_test = cv_data_update_test
)

saveRDS(kf_pred_update, file = "models/performance/results/cv_model_update_kfpred.rds")
beep("coin")
```

Fixed risk score model (without updating)
```{r}
kf_pred_noupdate <- kfold_manual(
  cvmodel = cv_model_noupdate,
  n_folds = length(unique(cv_data_noupdate_test$ID)),
  pca_features = significant_features,
  data_training = cv_data_noupdate_train,
  data_test = cv_data_noupdate_test
)

saveRDS(kf_pred_noupdate, file = "models/performance/results/cv_model_noupdate_kfpred.rds")
beep("coin")
```

"NUll model" risk score with demographics only
```{r}
kf_pred_demographics_only <- kfold_manual(
  cvmodel = cv_model_demographics_only,
  n_folds = length(unique(cv_data_update_test$ID)),
  pca_features = significant_features,
  data_training = cv_data_update_train,
  data_test = cv_data_update_test
)

saveRDS(kf_pred_demographics_only, file = "models/performance/results/cv_model_demographics_only_kfpred.rds")
beep("coin")
```


## Time-dependent AUROC

Load cross-validation results
```{r}
kf_pred_update <- readRDS("models/performance/results/cv_model_update_kfpred.rds")
kf_pred_noupdate <- readRDS("models/performance/results/cv_model_noupdate_kfpred.rds")
kf_pred_demographics_only <- readRDS("models/performance/results/cv_model_demographics_only_kfpred.rds")
```

Compute time-dependent AUROC for each risk score model
```{r, warning=F}
# Main risk score model (with updating)
AUC_stat_noupdate <- kf_pred_noupdate %>%
  group_by(ID, rel_day, event) %>%
  summarize(dismiss = mean(dismiss), stay = mean(stay), icu = mean(icu), .groups = "drop") %>%
  arrange(ID, rel_day) %>%
  group_by(ID) %>%
  mutate(tstart = rel_day - 1, tstop = ifelse(is.na(lead(tstart)), rel_day, lead(tstart))) %>%
  mutate(event = case_when(event == "stay" ~ 0, event == "dismiss" ~ 1, event == "icu" ~ 2)) %>%
  get_AUC_statistics(max_time = 7)

# Fixed risk score model (without updating)
AUC_stat_update <- kf_pred_update %>%
  group_by(ID, rel_day, event) %>%
  summarize(dismiss = mean(dismiss), stay = mean(stay), icu = mean(icu), .groups = "drop") %>%
  arrange(ID, rel_day) %>%
  group_by(ID) %>%
  mutate(tstart = rel_day - 1, tstop = ifelse(is.na(lead(tstart)), rel_day, lead(tstart))) %>%
  mutate(event = case_when(event == "stay" ~ 0, event == "dismiss" ~ 1, event == "icu" ~ 2)) %>%
  get_AUC_statistics(max_time = 7)

# "NUll model" risk score with demographics only
AUC_stat_demographics_only <- kf_pred_demographics_only %>%
  group_by(ID, rel_day, event) %>%
  summarize(dismiss = mean(dismiss), stay = mean(stay), icu = mean(icu), .groups = "drop") %>%
  arrange(ID, rel_day) %>%
  group_by(ID) %>%
  mutate(tstart = rel_day - 1, tstop = ifelse(is.na(lead(tstart)), rel_day, lead(tstart))) %>%
  mutate(event = case_when(event == "stay" ~ 0, event == "dismiss" ~ 1, event == "icu" ~ 2)) %>%
  get_AUC_statistics(max_time = 7)
```

Figure 5
```{r}
plot_auc_multi(AUC_stat_update[["dismiss"]], "Main (with daily update)", AUC_stat_noupdate[["dismiss"]], "Fixed (without update)")
ggsave("figures/performance_AUC.png", height = 9.26, width = 15, units = "cm", dpi = 300)
```

Figure 16 (Multimedia Appendix)
```{r,fig.eight=4.5,fig.width=7.29}
plot_auc_multi(AUC_stat_update[["dismiss"]], "Main (with daily update)", AUC_stat_demographics_only[["dismiss"]], "Only demographics", color2 = "#ff5c33ff", fill2 = "#ff5c3333")
ggsave("figures/performance_AUC_demographics.png", height = 9.26, width = 15, units = "cm", dpi = 300)
```
